qcircuit 2.5.3 Tutorial

### 2017-10-11 CO681-QuantumInformationProcessing/assignments/assignment_02.tex
problem 2

>>>>

\subsection{a}
\paragraph{}
We check the action of $H$ on $X$ and $Z$.
$$HXH^{\dagger} = \frac{1}{2} \begin{bmatrix} 1 & 1 \\ 1 &  -1 \end{bmatrix} \begin{bmatrix} 0 & 1 \\ 1 & 0 \end{bmatrix} \begin{bmatrix} 1 & 1 \\ 1 &  -1 \end{bmatrix} = \frac{1}{2} \begin{bmatrix} 1 & 1 \\ 1 &  -1 \end{bmatrix} \begin{bmatrix}1 & -1 \\ 1 & 1 \end{bmatrix} =\frac{1}{2} \begin{bmatrix} 2 & 0 \\ 0 & 2 \end{bmatrix} = I. $$
$$HZH^{\dagger} = \frac{1}{2} \begin{bmatrix} 1 & 1 \\ 1 &  -1 \end{bmatrix} \begin{bmatrix} 1 & 0 \\ 0 & -1 \end{bmatrix} \begin{bmatrix} 1 & 1 \\ 1 &  -1 \end{bmatrix} = \frac{1}{2} \begin{bmatrix} 1 & 1 \\ 1 &  -1 \end{bmatrix} \begin{bmatrix} 1 & 1 \\ -1 & 1\end{bmatrix} = \frac{1}{2} \begin{bmatrix} 0 & 2 \\ 2 & 0\end{bmatrix} = X.$$
Thus $H$ is in the Clifford group on $1$ qubit.
\subsection{b}
\paragraph{}
Similar to the previous subsection we consider the action of $T$ on $X$ and $Z$.
$$TXT^{\dagger} = \begin{bmatrix} 1 & 0 \\ 0& \zeta_8 \end{bmatrix}\begin{bmatrix} 0 & 1 \\ 1 & 0 \end{bmatrix}\begin{bmatrix} 1 & 0 \\ 0& \zeta_8^* \end{bmatrix} = \begin{bmatrix} 1 & 0 \\ 0& \zeta_8 \end{bmatrix}\begin{bmatrix} 0 & \zeta^*_8 \\ 1 & 0 \end{bmatrix} = \begin{bmatrix} 0 & \zeta_8^* \\ \zeta_8 & 0 \end{bmatrix} = -e^{\frac{3}{4}i\pi}X.$$
$$TZT^{\dagger} = \begin{bmatrix} 1 & 0 \\ 0& \zeta_8 \end{bmatrix}\begin{bmatrix} 1 & 0 \\ 0 & -1 \end{bmatrix}\begin{bmatrix} 1 & 0 \\ 0& \zeta_8^* \end{bmatrix} = \begin{bmatrix} 1 & 0 \\ 0& \zeta_8 \end{bmatrix}\begin{bmatrix} 1 & 0 \\ 0 & -\zeta^*_8\end{bmatrix} = \begin{bmatrix} 1 & 0 \\ 0 & -1\end{bmatrix} = Z.$$
Thus $T$ is in the Clifford group on $1$ qubit.
\subsection{c}
\paragraph{}
$$SXS^\dagger = \begin{bmatrix} 1 & 0 \\ 0 & i \end{bmatrix} \begin{bmatrix} 0 & 1 \\ 1 & 0 \end{bmatrix} \begin{bmatrix} 1 & 0\\ 0 & -i \end{bmatrix} = \begin{bmatrix} 1 & 0 \\ 0 & i \end{bmatrix} \begin{bmatrix} 0 & -i\\ 1 & 0 \end{bmatrix} =\begin{bmatrix} 0 & -i \\ i & 0 \end{bmatrix} = Y.$$
$$SZS^\dagger = \begin{bmatrix} 1 & 0 \\ 0 & i \end{bmatrix}\begin{bmatrix} 1 & 0 \\ 0 & -1\end{bmatrix} \begin{bmatrix} 1 & 0 \\0 & -i \end{bmatrix} = \begin{bmatrix} 1 & 0 \\ 0 & i \end{bmatrix}\begin{bmatrix} 1 & 0 \\ 0 & i \end{bmatrix} = Z.$$
Thus $S$ is in the Clifford group on $1$ qubit.
\subsection{d}
\paragraph{}
Let $C$ be the unitary matrix representing the controlled-not gate. We consider the action of $C$ on $I \otimes X$, $X\otimes I$, $I \otimes Z$. and $Z \otimes I$.
$$C (I\otimes X) C^\dagger = \begin{bmatrix} 1 & 0 & 0 & 0 \\ 0 & 1 & 0 & 0 \\ 0 & 0 & 0 &1 \\ 0& 0& 1 & 0 \end{bmatrix}\begin{bmatrix}0 & 1 & 0 & 0 \\ 1 & 0 & 0 &0 \\ 0 & 0 &0 & 1 \\ 0 & 0 & 1 & 0 \end{bmatrix} \begin{bmatrix} 1 & 0 & 0 & 0 \\ 0 & 1 & 0 & 0 \\ 0 & 0 & 0 &1 \\ 0& 0& 1 & 0 \end{bmatrix} = \begin{bmatrix} 0 & 1& 0& 0\\ 1 & 0  & 0& 0\\ 0 & 0 & 1 & 0\\ 0 & 0 & 0 & 1\end{bmatrix} \begin{bmatrix} 1 & 0 & 0 & 0 \\ 0 & 1 & 0 & 0 \\ 0 & 0 & 0 &1 \\ 0& 0& 1 & 0 \end{bmatrix} = I \otimes X.$$
$$ C (X \otimes I) C^\dagger =  \begin{bmatrix} 1 & 0 & 0 & 0 \\ 0 & 1 & 0 & 0 \\ 0 & 0 & 0 &1 \\ 0& 0& 1 & 0 \end{bmatrix} \begin{bmatrix}  0 & 0 & 1 & 0 \\ 0 & 0 & 0 & 1 \\ 1 & 0 & 0 & 0 \\ 0 &1 & 0 &0\end{bmatrix} \begin{bmatrix} 1 & 0 & 0 & 0 \\ 0 & 1 & 0 & 0 \\ 0 & 0 & 0 &1 \\ 0& 0& 1 & 0 \end{bmatrix}  =  \begin{bmatrix} 0 & 0 &1 & 0 \\ 0 & 0 & 0 & 1 \\ 0 & 1 & 0 & 0 \\ 1 & 0 & 0 &0\end{bmatrix} \begin{bmatrix} 1 & 0 & 0 & 0 \\ 0 & 1 & 0 & 0 \\ 0 & 0 & 0 &1 \\ 0& 0& 1 & 0 \end{bmatrix} = X \otimes X$$
\begin{align*}C (I \otimes Z) C^\dagger &=  \begin{bmatrix} 1 & 0 & 0 & 0 \\ 0 & 1 & 0 & 0 \\ 0 & 0 & 0 &1 \\ 0& 0& 1 & 0 \end{bmatrix}\begin{bmatrix}1 &0 &0 &0 \\ 0 & -1 & 0 &0 \\ 0 & 0 &1 & 0\\ 0& 0& 0 & -1 \end{bmatrix} \begin{bmatrix} 1 & 0 & 0 & 0 \\ 0 & 1 & 0 & 0 \\ 0 & 0 & 0 &1 \\ 0& 0& 1 & 0 \end{bmatrix}\\
 &=  \begin{bmatrix} 1 & 0 & 0 &0 \\ 0 & -1 & 0 &0 \\ 0 & 0 & 0 & -1 \\ 0 &0 &1 & 0\end{bmatrix}\begin{bmatrix} 1 & 0 & 0 & 0 \\ 0 & 1 & 0 & 0 \\ 0 & 0 & 0 &1 \\ 0& 0& 1 & 0 \end{bmatrix}\\ &= \begin{bmatrix} 1 & 0 &0 &0 \\ 0 & -1 &0 &0 \\ 0 &0 &-1 & 0\\ 0 &0 &0 &1\end{bmatrix} = Z \otimes Z.\end{align*}
 \begin{align*}
 C(Z\otimes I)C^\dagger &=  \begin{bmatrix} 1 & 0 & 0 & 0 \\ 0 & 1 & 0 & 0 \\ 0 & 0 & 0 &1 \\ 0& 0& 1 & 0 \end{bmatrix} \begin{bmatrix}1 &0 &0 &0 \\0 & 1 & 0 &0 \\ 0 &0 &-1 & 0\\ 0 &0 &0 & -1 \end{bmatrix}  \begin{bmatrix} 1 & 0 & 0 & 0 \\ 0 & 1 & 0 & 0 \\ 0 & 0 & 0 &1 \\ 0& 0& 1 & 0 \end{bmatrix}\\
 &= \begin{bmatrix} 1& 0 & 0 &0 \\0 &1 & 0 &0 \\ 0 & 0 &0 &-1 \\ 0 & 0 & -1 &0\end{bmatrix} \begin{bmatrix} 1 & 0 & 0 & 0 \\ 0 & 1 & 0 & 0 \\ 0 & 0 & 0 &1 \\ 0& 0& 1 & 0 \end{bmatrix}\\
 &= \begin{bmatrix} 1 & 0 &0 &0 \\ 0 &1 & 0 &0\\ 0 & 0 &-1 & 0 \\ 0 & 0 & 0 & -1 \end{bmatrix} = Z\otimes I.
 \end{align*}
 Thus $C$ is in the $2$-qubit Clifford group.
 \subsection{e}
 Let $A$ be the Toffoli gate (CCNOT). For the sake of space we'll abbreviate the calculations. We have that
 $$A(I\otimes X\otimes I) A^\dagger = \begin{bmatrix} X\otimes I & 0 \\ 0 & Z\otimes Z\end{bmatrix}.$$
 This can't be written as a tensor product $\alpha V_1\otimes V_2 \otimes V_3$ with $\alpha$ scalar and $V_1,V_2, V_3 \in \{I,X,Y,Z\}$. To see this observe that if we take $V_1 \in  \{X, Y\}$ then the first $4x4$ diagonal block of $V_1\otimes V_2 \otimes V_3$ will be $0$, contradicting that it should be $X \otimes I$, and if we take $V_1 \in \{I,Z\}$ then the first $4x4$ diagonal block will be $\pm1$ times the second, but $X\otimes I \neq \pm Z\otimes Z$. Therefore no such tensor product expression for $A(I\otimes X \otimes I)A^\dagger$ is possible, and hence the Toffoli gate is not in the $3$-qubit Clifford group. $\blacksquare$
